Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-10-26 manuscript/climate_sensitivity_draft.Rmd
Updated file names

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

@@

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

\noindent\hrulefill 
```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

@@

Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#) and then used these to estimate bark thickness based on DBH.

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013]. Species-specific height allometries were developed (Table S1). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.

@@

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/species_table.csv")

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/univariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  kable_styling()
```

@@

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

>>>>

**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

@@

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

\noindent\hrulefill 
```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

@@

Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH. Additional species were used here than used in the final analysis to establish base averages. (**do we want to keep this or take out juni/fram from table?**)

Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013]. Species-specific height allometries were developed (Table S2). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.

@@

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/table3_species_table.csv")

@@

```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(univar_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```

```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(multivar_table, booktabs = TRUE, caption = "**Table 5. Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  kable_styling()
```

@@

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
